New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2024-09-17 jso.qmd
minor corrections

where $4$ is a constant scaling factor, $-\min_d g'(d)$ represents the 
largest gradient of $-g$ and $\arg \min_{d} g'(d)$ represents the projection distance at which this largest gradient is attained.

@@

It is expected that for a PP index with high squint angle, the
optimization (@eq-optimization) should make substantial progress early
on. Conversely, for a PP index with low squint angle, it might take a
long while for the optimization to make substantial progress, as the
candidate projections would need to be very close to the optimal one for
the structure of the index function to be visible enough to be amenable
to efficient optimization. This observation suggests that the extreme
values of $g'$ (the ones for which $g''=0$, assuming that $g$ is twice
differentiable), and the projection distances for which these values are
attained, are crucial in the mathematical definition of squintability.

<!-- Noticing that $f$ is expected to be a decreasing function, the -->
<!-- squintability metric is proposed as follows: -->

<!-- $$ -->
<!-- \varsigma:=-4\min_{x}f'(x)\times\underset{x}{\arg\min}f'(x)\ . -->
<!-- $$ {#eq-squintability} -->

<!-- $-\min_{x}f'(x)$ represents the largest value of the gradient of $-f$ -->
<!-- and $\underset{x}{\arg\min}f'(x)$ represents the projection distance at -->
<!-- which this largest gradient is attained.  -->

To compute the squintability metric \eqref{eq-squintability} in practice,
several approaches are possible. The first one is to propose a
parametric model for $g$, and use it to obtain an explicit formula for
$\varsigma$. Numerical experiments suggest a scaled sigmoid shape as
described below. Define

$$
\ell(x):=\frac{1}{1+\exp(\theta_{3}(x-\theta_{2}))}\ ,
$$ {#eq-logistic}

which is a decreasing logistic function depending on two parameters
$\theta_2$ and $\theta_3$, such that $\ell(\theta_{2})=\frac{1}{2}$.
Then, define

$$
g(x)=(\theta_{1}-\theta_{4})\frac{\ell(x)-\ell(x_{\max})}{\ell(0)-\ell(x_{\max})}+\theta_{4}\ ,
$$ {#eq-parametric}

which depends on three additional parameters, $\theta_1$, $\theta_2$,
and $x_{\max}$, such that $g(0)=\theta_1$ and $g(x_{\max})=\theta_4$.
Under the parametric model (@eq-parametric), the squintability metric
\eqref{eq-squintability} can be shown to be equal to

$$
\varsigma=\frac{(\theta_{1}-\theta_{4})\theta_{2}\theta_{3}}{\ell(0)-\ell(x_{\max})}\ .
$$ {#eq-squintability-parametric}

@@

To compute squintability, 50 random bases are sampled and interpolated
to the optimal basis with a step size of 0.005. Index values and
projection distances are calculated for these interpolated bases and the
index values are averaged with a bin width of 0.005. A four-parameter
scaled logistic function is fitted to the index values against
projection distances, estimated by non-linear least squares. The
squintability measure is then calculated as
(@eq-squintability-parametric).

@@

```{r}
#| label = fig-proportion,
#| fig.width = 8,
#| fig.height = 3.5,
#| fig.align = "center",
#| fig.cap = "The proportion of simulations that reach near-optimal index values in the pipe-finding problem using the holes index. The proportion is calculated based on the number of simulations, out of 50, that achieve an index value within 0.05 of the best-performing simulation. To quantify uncertainty, Bootstrap samples with 500 are generated. The thin lines represent the proportion for each of the 500 bootstrap replicates, while the thicker lines represent the mean of these bootstrap samples, connnected by lines. As the dimensionality increases, the proportion of simulations reaching the optimal index value increases."
pipe_sim_best <- sim_pipe_run_best |> 
  group_by(n_jellies, max_tries, d) |>
  summarise(I_best = max(I_max))

@@

@tbl-mod-output presents the results from fitting a logistic regression model using proportion of success as the response and smoothness, squintability, dimension, running time, number of jellyfish and maximum number of tries, as predictors. Long runtime is a binary variable, where greater than 20 seconds is labelled 1. The fit suggests that JSO success rate is only affected by squintability, dimension and number of jellies. Interestingly, the smoothness of the index is not a problem for the JSO - this is promising! Running the optimisation longer and allowing more tries also does not affect success rate.  An increase in squintability of one unit increases the success rate by `r round(exp(broom::tidy(mod1)$estimate[3]) * 100, 0)`%. As dimension increases by one the success rate halves. Increasing the number of jellies by 10, increases the succes rate by `r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 0)`%.

>>>>

where $4$ is a constant scaling factor, $-\min_d g'(d)$ represents the 
largest gradient of $-g$ and $\arg \min_{d} g'(d)$ represents the projection distance at which this largest gradient is attained. If the set $\mathcal{D}:=\arg \min_{d} g'(d)$ contains more than one value $d$, the largest value $\max(\mathcal{D})$ is used in equation \eqref{eq-squintability}.

@@

It is expected that for a PP index with high squint angle, the
optimization (@eq-optimization) should make substantial progress early
on. Conversely, for a PP index with low squint angle, it might take a
long while for the optimization to make substantial progress, as the
candidate projections would need to be very close to the optimal one for
the structure of the index function to be visible enough to be amenable
to efficient optimization. This observation suggests that the extreme
values of $g'$ and the projection distances for which these values are
attained, are crucial in the mathematical definition of squintability.

To compute the squintability metric \eqref{eq-squintability} in practice,
several approaches are possible. The first one is to propose a
parametric model for $g$, and use it to obtain an explicit formula for
$\varsigma$. Numerical experiments suggest a scaled sigmoid shape as
described below. Define
$$
\ell(x):=\frac{1}{1+\exp(\theta_{3}(x-\theta_{2}))}\ ,
$$ {#eq-logistic}

which is a decreasing logistic function depending on two parameters
$\theta_2$ and $\theta_3$, such that $\ell(\theta_{2})=\frac{1}{2}$.
Then, define
$$
g(x)=(\theta_{1}-\theta_{4})\frac{\ell(x)-\ell(x_{\max})}{\ell(0)-\ell(x_{\max})}+\theta_{4}\ ,
$$ {#eq-parametric}

which depends on three additional parameters, $\theta_1$, $\theta_2$,
and $x_{\max}$, such that $g(0)=\theta_1$ and $g(x_{\max})=\theta_4$.
Under the parametric model (@eq-parametric), the squintability metric
\eqref{eq-squintability} can be shown to be equal to
$$
\varsigma=\frac{(\theta_{1}-\theta_{4})\theta_{2}\theta_{3}}{\ell(0)-\ell(x_{\max})}\ .
$$ {#eq-squintability-parametric}

@@

To compute squintability, 50 random bases are sampled and interpolated
to the optimal basis with a step size of 0.005. Index values and
projection distances are calculated for these interpolated bases and the
index values are averaged with a bin width of 0.005. The four-parameter
scaled logistic function (@eq-parametric) is fitted to the index values against
projection distances, estimated by non-linear least squares. The
squintability measure is then calculated as
(@eq-squintability-parametric).

@@

```{r}
#| label = fig-proportion,
#| fig.width = 8,
#| fig.height = 3.5,
#| fig.align = "center",
#| fig.cap = "The proportion of simulations that reach near-optimal index values in the pipe-finding problem using the holes index. The proportion is calculated based on the number of simulations, out of 50, that achieve an index value within 0.05 of the best-performing simulation. To quantify uncertainty, 500 bootstrap samples are generated. The thin segments represent the proportion for each of the 500 bootstrap replicates, while the thicker segments represent the mean of these bootstrap samples, connnected by lines. As the dimensionality increases, the proportion of simulations reaching the optimal index value decreases."
pipe_sim_best <- sim_pipe_run_best |> 
  group_by(n_jellies, max_tries, d) |>
  summarise(I_best = max(I_max))

@@

@tbl-mod-output presents the results from fitting a logistic regression model using the proportion of success as the response and smoothness, squintability, dimension, running time, number of jellyfish and maximum number of tries, as predictors. Long runtime is a binary variable, where greater than 20 seconds is labelled 1. The fit suggests that JSO success rate is only affected by squintability, dimension and number of jellies. Interestingly, the smoothness of the index is not a problem for the JSO - this is promising! Running the optimisation longer and allowing more tries also does not affect success rate.  An increase in squintability of one unit increases the success rate by `r round(exp(broom::tidy(mod1)$estimate[3]) * 100, 0)`%. As dimension increases by one the success rate halves. Increasing the number of jellies by 10, increases the succes rate by `r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 0)`%.

### 2024-09-17 jso.tex
minor corrections

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. Nicolas
Langrené. Email: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}. Jessica
Wai Yin
Leung. Email: \href{mailto:Jessica.Leung@monash.edu}{\nolinkurl{Jessica.Leung@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Austin, United
States\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash
University, Melbourne, Australia\\ \textsuperscript{3} Department of
Mathematical Sciences, Guangdong Provincial Key Laboratory of
Interdisciplinary Research and Application for Data Science,
BNU-HKBU~United~International~College, Zhuhai, China
\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criterion function known as the PP index, to explore high-dimensional
data by revealing interesting projections. Optimisation of some PP
indexes can be non-trivial, if they are non-smooth functions, or the
optimum has a small ``squint angle'', detectable only from close
proximity. To address these challenges, this study investigates the
performance of a recently introduced swarm-based algorithm, Jellyfish
Search Optimiser (JSO), for optimising PP indexes. The performance of
JSO for visualising data is evaluated across various hyper-parameter
settings and compared with existing optimisers. Additionally, methods
for calculating the smoothness and squintability properties of the PP
index are proposed. They are used to assess the optimiser performance in
the presence of PP index complexities. A simulation study illustrates
the use of these performance metrics to compare the JSO with existing
optimisation methods available for the guided tour. The JSO algorithm
has been implemented in the R package, \texttt{tourr}, and functions to
calculate smoothness and squintability are available in the
\texttt{ferrn} package.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
projection pursuit\sep jellyfish search optimiser
(JSO)\sep optimisation\sep grand tour\sep high-dimensional data\sep 
exploratory data analysis
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, enhanced]}{\end{tcolorbox}}\fi

@@

It is expected that for a PP index with high squint angle, the
optimization (\ref{eq-optimization}) should make substantial progress
early on. Conversely, for a PP index with low squint angle, it might
take a long while for the optimization to make substantial progress, as
the candidate projections would need to be very close to the optimal one
for the structure of the index function to be visible enough to be
amenable to efficient optimization. This observation suggests that the
extreme values of \(g'\) (the ones for which \(g''=0\), assuming that
\(g\) is twice differentiable), and the projection distances for which
these values are attained, are crucial in the mathematical definition of
squintability.

To compute the squintability metric \eqref{eq-squintability} in
practice, several approaches are possible. The first one is to propose a
parametric model for \(g\), and use it to obtain an explicit formula for
\(\varsigma\). Numerical experiments suggest a scaled sigmoid shape as
described below. Define

\begin{equation}\protect\hypertarget{eq-logistic}{}{
\ell(x):=\frac{1}{1+\exp(\theta_{3}(x-\theta_{2}))}\ ,
}\label{eq-logistic}\end{equation}

which is a decreasing logistic function depending on two parameters
\(\theta_2\) and \(\theta_3\), such that
\(\ell(\theta_{2})=\frac{1}{2}\). Then, define

\begin{equation}\protect\hypertarget{eq-parametric}{}{
g(x)=(\theta_{1}-\theta_{4})\frac{\ell(x)-\ell(x_{\max})}{\ell(0)-\ell(x_{\max})}+\theta_{4}\ ,
}\label{eq-parametric}\end{equation}

which depends on three additional parameters, \(\theta_1\),
\(\theta_2\), and \(x_{\max}\), such that \(g(0)=\theta_1\) and
\(g(x_{\max})=\theta_4\). Under the parametric model
(\ref{eq-parametric}), the squintability metric \eqref{eq-squintability}
can be shown to be equal to

\begin{equation}\protect\hypertarget{eq-squintability-parametric}{}{
\varsigma=\frac{(\theta_{1}-\theta_{4})\theta_{2}\theta_{3}}{\ell(0)-\ell(x_{\max})}\ .
}\label{eq-squintability-parametric}\end{equation}

@@

\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, leftrule=.75mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, rightrule=.15mm, opacityback=0, breakable, coltitle=black, colframe=quarto-callout-note-color-frame, left=2mm, toprule=.15mm, bottomrule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, opacitybacktitle=0.6, colback=white]

@@

To compute squintability, 50 random bases are sampled and interpolated
to the optimal basis with a step size of 0.005. Index values and
projection distances are calculated for these interpolated bases and the
index values are averaged with a bin width of 0.005. A four-parameter
scaled logistic function is fitted to the index values against
projection distances, estimated by non-linear least squares. The
squintability measure is then calculated as
(\ref{eq-squintability-parametric}).

@@

\caption{\label{fig-proportion}The proportion of simulations that reach
near-optimal index values in the pipe-finding problem using the holes
index. The proportion is calculated based on the number of simulations,
out of 50, that achieve an index value within 0.05 of the
best-performing simulation. To quantify uncertainty, Bootstrap samples
with 500 are generated. The thin lines represent the proportion for each
of the 500 bootstrap replicates, while the thicker lines represent the
mean of these bootstrap samples, connnected by lines. As the
dimensionality increases, the proportion of simulations reaching the
optimal index value increases.}

@@

Table~\ref{tbl-mod-output} presents the results from fitting a logistic
regression model using proportion of success as the response and
smoothness, squintability, dimension, running time, number of jellyfish
and maximum number of tries, as predictors. Long runtime is a binary
variable, where greater than 20 seconds is labelled 1. The fit suggests
that JSO success rate is only affected by squintability, dimension and
number of jellies. Interestingly, the smoothness of the index is not a
problem for the JSO - this is promising! Running the optimisation longer
and allowing more tries also does not affect success rate. An increase
in squintability of one unit increases the success rate by 785\%. As
dimension increases by one the success rate halves. Increasing the
number of jellies by 10, increases the succes rate by 24\%.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. Nicolas
Langrené. Email: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}. Jessica
Wai Yin
Leung. Email: \href{mailto:Jessica.Leung@monash.edu}{\nolinkurl{Jessica.Leung@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Austin, United
States\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash
University, Melbourne, Australia\\ \textsuperscript{3} Department of
Mathematical Sciences, Guangdong Provincial Key Laboratory of
Interdisciplinary Research and Application for Data Science,
BNU-HKBU~United~International~College, Zhuhai, China
\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criterion function known as the PP index, to explore high-dimensional
data by revealing interesting projections. Optimisation of some PP
indexes can be non-trivial, if they are non-smooth functions, or the
optimum has a small ``squint angle'', detectable only from close
proximity. To address these challenges, this study investigates the
performance of a recently introduced swarm-based algorithm, Jellyfish
Search Optimiser (JSO), for optimising PP indexes. The performance of
JSO for visualising data is evaluated across various hyper-parameter
settings and compared with existing optimisers. Additionally, methods
for calculating the smoothness and squintability properties of the PP
index are proposed. They are used to assess the optimiser performance in
the presence of PP index complexities. A simulation study illustrates
the use of these performance metrics to compare the JSO with existing
optimisation methods available for the guided tour. The JSO algorithm
has been implemented in the R package, \texttt{tourr}, and functions to
calculate smoothness and squintability are available in the
\texttt{ferrn} package.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
projection pursuit\sep jellyfish search optimiser
(JSO)\sep optimisation\sep grand tour\sep high-dimensional data\sep 
exploratory data analysis
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

It is expected that for a PP index with high squint angle, the
optimization (\ref{eq-optimization}) should make substantial progress
early on. Conversely, for a PP index with low squint angle, it might
take a long while for the optimization to make substantial progress, as
the candidate projections would need to be very close to the optimal one
for the structure of the index function to be visible enough to be
amenable to efficient optimization. This observation suggests that the
extreme values of \(g'\) and the projection distances for which these
values are attained, are crucial in the mathematical definition of
squintability.

To compute the squintability metric \eqref{eq-squintability} in
practice, several approaches are possible. The first one is to propose a
parametric model for \(g\), and use it to obtain an explicit formula for
\(\varsigma\). Numerical experiments suggest a scaled sigmoid shape as
described below. Define
\begin{equation}\protect\hypertarget{eq-logistic}{}{
\ell(x):=\frac{1}{1+\exp(\theta_{3}(x-\theta_{2}))}\ ,
}\label{eq-logistic}\end{equation}

which is a decreasing logistic function depending on two parameters
\(\theta_2\) and \(\theta_3\), such that
\(\ell(\theta_{2})=\frac{1}{2}\). Then, define
\begin{equation}\protect\hypertarget{eq-parametric}{}{
g(x)=(\theta_{1}-\theta_{4})\frac{\ell(x)-\ell(x_{\max})}{\ell(0)-\ell(x_{\max})}+\theta_{4}\ ,
}\label{eq-parametric}\end{equation}

which depends on three additional parameters, \(\theta_1\),
\(\theta_2\), and \(x_{\max}\), such that \(g(0)=\theta_1\) and
\(g(x_{\max})=\theta_4\). Under the parametric model
(\ref{eq-parametric}), the squintability metric \eqref{eq-squintability}
can be shown to be equal to
\begin{equation}\protect\hypertarget{eq-squintability-parametric}{}{
\varsigma=\frac{(\theta_{1}-\theta_{4})\theta_{2}\theta_{3}}{\ell(0)-\ell(x_{\max})}\ .
}\label{eq-squintability-parametric}\end{equation}

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, breakable, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, toprule=.15mm, colback=white, left=2mm, opacityback=0, coltitle=black, opacitybacktitle=0.6, arc=.35mm, rightrule=.15mm]

@@

To compute squintability, 50 random bases are sampled and interpolated
to the optimal basis with a step size of 0.005. Index values and
projection distances are calculated for these interpolated bases and the
index values are averaged with a bin width of 0.005. The four-parameter
scaled logistic function (\ref{eq-parametric}) is fitted to the index
values against projection distances, estimated by non-linear least
squares. The squintability measure is then calculated as
(\ref{eq-squintability-parametric}).

@@

\caption{\label{fig-proportion}The proportion of simulations that reach
near-optimal index values in the pipe-finding problem using the holes
index. The proportion is calculated based on the number of simulations,
out of 50, that achieve an index value within 0.05 of the
best-performing simulation. To quantify uncertainty, 500 bootstrap
samples are generated. The thin segments represent the proportion for
each of the 500 bootstrap replicates, while the thicker segments
represent the mean of these bootstrap samples, connnected by lines. As
the dimensionality increases, the proportion of simulations reaching the
optimal index value decreases.}

@@

Table~\ref{tbl-mod-output} presents the results from fitting a logistic
regression model using the proportion of success as the response and
smoothness, squintability, dimension, running time, number of jellyfish
and maximum number of tries, as predictors. Long runtime is a binary
variable, where greater than 20 seconds is labelled 1. The fit suggests
that JSO success rate is only affected by squintability, dimension and
number of jellies. Interestingly, the smoothness of the index is not a
problem for the JSO - this is promising! Running the optimisation longer
and allowing more tries also does not affect success rate. An increase
in squintability of one unit increases the success rate by 785\%. As
dimension increases by one the success rate halves. Increasing the
number of jellies by 10, increases the succes rate by 24\%.
